\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-19 k-planar.tex
Trying a restructuring

The following theorem is the main contribution of the paper. It says that if a graph class $\mathcal{G}$ has a product structure theorem, then the class of graphs obtained by applying a shortcut system to graphs in $\mathcal{G}$ also has a product structure theorem.

@@

Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} and \cref{AddDummy} imply \cref{kPlanarProduct} with $K_{6(k^3+3k)}$ instead of $K_{18k^2+48k+30}$. Some further observations presented in \cref{sec-k-planar} lead to the improved result.

\cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Here is one example. The \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\SS$ for some $(k,2k\Delta^{k})$-shortcut system $\SS$; see \cref{PowerShortcut}. Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} then imply:

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is contained in $H\boxtimes P\boxtimes K_{6k^2(k^2+3)\Delta^{k}}$ for some graph $H$ of treewidth at most $\binom{k+3}{3}-1$ and some path $P$.
\end{thm}

\cref{Examples} presents further examples of graph classes that can be constructed using shortcut systems, including certain types of map graphs, string graphs, and $k$-nearest neighbour graphs. \cref{ShortcutProduct} implies product structure theorems for each of these classes. All of the above-mentioned applications also hold for these examples.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Shortcut Systems}
\label{Structure}

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, an $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

We are now ready to prove our main result, which we restate here for convenience:

\mmg*

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Framed Graphs}

\note{PM}{I don't like this structure anymore.  The proof of \cref{kPlanarProduct} ($k$-planar) refers to the innards of the proof of \cref{ShortcutPartition} ($(k,d)$-shortcuts).  The proofs should be close to each other.

Let's consider something like this:
\begin{compactenum}
  \item Introduction
  \item Shortcut systems and $k$-planar graphs
  \item All graphs obtained from shortcut systems or $k$-planar graphs
  \item $d$-framed graphs, $1$-planar graphs, and map graphs
  \item All applications (colouring, labelling, etc).
\end{compactenum}
% (Possibly with the order of 2 and 3 reversed.)
}

@@

\subsection{$k$-Planar Graphs}

\note{DW}{These definitions should appear earlier, certainly before the definition of framed graph, which should be defined as an ``embedded graph''. One option is to put this whole paragraph as a footnote in the introduction where we first mention $k$-planar graphs. The experienced reader can skip this paragraph, which says to me it should be a footnote.}
We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \defin{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior. The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$. A \defin{crossing} in an embedded (multi)graph $G$ is an unordered pair of distinct edges $\{vw,xy\}\subseteq E(G)$ whose interiors have a point in common. Without any loss of generality we may assume that the interiors of any two edges of $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$ then we say that $vw$ and $xy$ \defin{cross at} $p$.

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph is contained in $H\boxtimes P\boxtimes K_7$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

@@

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma. Similar results appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Examples}
\label{Examples}

@@

@@

%%%%%%%
\subsection{String Graphs}

\begin{thm}
\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph:
\begin{compactitem}
	\item is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$.
\end{compactitem}
\end{thm}

Our results also give bounds on the non-repetitive chromatic number and the $p$-centered chromatic number of $(g,\delta)$-string graphs, but the bounds are weak, since such graphs $G$ have maximum degree at most $2\delta$, implying that $\pi(G) \leq (4+o(1))\delta^2$ and $\chi_p(G)\le p(64\delta)^2$ by results of \citet{DJKW16} and \citet{DFMS21}, respectively.

%%%%%%%%%%
\subsection{Powers of Bounded Degree Graphs}
\label{Powers}

Theorems~\ref{PlanarProduct}(b), \ref{GenusProduct}(b) and \cref{ShortcutProduct} and  \cref{DrawG,qn,p-centered,non-repetitive,PowerShortcut} imply the following result, which with $g=0$ implies  \cref{kPowerBasic} in the introduction.

\begin{thm}
\label{PowerGenus}
For integers $g\geq 0$ and $k,\Delta\geq 1$, let $\ell:= \max\{2g,3\} (2k^4+6k^2) \Delta^{k}$ and $t:= \binom{k+3}{3}-1$. Then for every graph $G$ of Euler genus $g$ and maximum degree $\Delta$,
\begin{compactitem}
\item $G^k$ is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$
\item $G^k$ is $(g, 2k(k+1)\Delta^{k} )$-planar,
\item $G^k$ has queue-number $\qn(G^k) \leq 3 \ell \cdot 2^t - \ceil{\tfrac{3}{2}\ell}$.
\item $G^k$ has  $p$-centered chromatic number $\chi_p(G^k) \leq \ell (p+1)\,  \binom{p+t}{t}$,
\item $G^k$ has  non-repetitive chromatic number $ \pi(G^k) \leq \ell \, 4^{t+1}$.
\end{compactitem}
\end{thm}

% \note{DW}{\cref{PowerGenus} has been updated}

This result is the first constant upper bound on the queue-number of bounded powers of graphs with bounded degree and bounded Euler genus.  For every graph $G$, since $G^k$ has maximum degree at most $\Delta^k$, a result of \citet{DJKW16} implies that $\pi(G^k) \leq (1+o(1))\Delta^{2k}$. \cref{PowerGenus} improves upon this bound when $k,g\ll\Delta$.  Similarly, a result of \citet{DFMS21} implies that $\chi_p(G^k)\le 1024p\Delta^{2k}$ and \cref{PowerGenus} improves upon this bound when $p,k,g\ll\Delta$.

\cref{qn,p-centered,non-repetitive,MinorFreeDegree,PowerProduct} imply the following analogous result for powers of graphs in any minor-closed class with bounded maximum degree.

\begin{thm}
\label{PowerMinor}
For every graph $X$ there exists an integer $c$ such that for all integers $k,\Delta\geq 1$, if $t:= 2k\Delta^{k}(k^3+3k)\binom{k+c\Delta}{c\Delta}-1$ and $G$ is an $X$-minor-free graph with maximum degree $\Delta$, then:
\begin{compactitem}
	\item $G^k$ is contained in $H\boxtimes P$ for some graph $H$ with treewidth $t$ and for some path $P$.
	\item $G^k$ has queue-number at most $3\cdot 2^t-2$,
	\item $G^k$ has $p$-centered chromatic number at most $(p+1)\binom{p+t}{t}$.
\end{compactitem}
\end{thm}

\subsection{$k$-Nearest-Neighbour Graphs}

@@

>>>>

The following theorem is one of the main contributions of the paper. It says that if a graph class $\mathcal{G}$ has a product structure theorem, then the class of graphs obtained by applying a shortcut system to graphs in $\mathcal{G}$ also has a product structure theorem.

@@

Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} and \cref{AddDummy} imply \cref{kPlanarProduct} with $K_{6(k^3+3k)}$ instead of $K_{18k^2+48k+30}$. Some further observations presented in \cref{sec-k-planar} lead to the improved result.  \cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Some examples are explored in \cref{first_example,mid_example,last_example}.

% Here is one example. The \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\SS$ for some $(k,2k\Delta^{k})$-shortcut system $\SS$; see \cref{PowerShortcut}. Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} then imply:

% \begin{thm}
% \label{kPowerBasic}
% For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is contained in $H\boxtimes P\boxtimes K_{6k^2(k^2+3)\Delta^{k}}$ for some graph $H$ of treewidth at most $\binom{k+3}{3}-1$ and some path $P$.
% \end{thm}

% \cref{Examples} presents further examples of graph classes that can be constructed using shortcut systems, including certain types of map graphs, string graphs, and $k$-nearest neighbour graphs. \cref{ShortcutProduct} implies product structure theorems for each of these classes. All of the above-mentioned applications also hold for these examples.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\boldmath Shortcut Systems and $k$-Planar Graphs}
\label{Structure}

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}, and its specialization to $k$-planar graphs \cref{kPlanarProduct}.

\subsection{Product Structure for Shortcut Systems}

\cref{ShortcutPartition} shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, an $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

We are now ready to prove our main result.
% , which we restate here for convenience:

% \mmg*

\begin{proof}[Proof of \cref{ShortcutPartition}]
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.

@@

\subsection{\boldmath Product Structure for $k$-Planar Graphs}

% \note{DW}{These definitions should appear earlier, certainly before the definition of framed graph, which should be defined as an ``embedded graph''. One option is to put this whole paragraph as a footnote in the introduction where we first mention $k$-planar graphs. The experienced reader can skip this paragraph, which says to me it should be a footnote.}

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \defin{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior. The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$. A \defin{crossing} in an embedded (multi)graph $G$ is an unordered pair of distinct edges $\{vw,xy\}\subseteq E(G)$ whose interiors have a point in common. Without any loss of generality we may assume that the interiors of any two edges of $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$ then we say that $vw$ and $xy$ \defin{cross at} $p$.

% \subsection{Examples}

%%%%%%%
\subsection{String Graphs}
\label{first_example}

\begin{thm}\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
\end{thm}

% \begin{thm}
% \label{StringPartition}
% For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
% Then  every $(g,\delta)$-string graph:
% \begin{compactitem}
% 	\item is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
% 	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$.
% \end{compactitem}
% \end{thm}

% Our results also give bounds on the non-repetitive chromatic number and the $p$-centered chromatic number of $(g,\delta)$-string graphs, but the bounds are weak, since such graphs $G$ have maximum degree at most $2\delta$, implying that $\pi(G) \leq (4+o(1))\delta^2$ and $\chi_p(G)\le p(64\delta)^2$ by results of \citet{DJKW16} and \citet{DFMS21}, respectively.

%%%%%%%%%%
\subsection{Powers of Bounded Degree Graphs}
\label{Powers}
\label{mid_example}

% Theorems~\ref{PlanarProduct}(b), \ref{GenusProduct}(b) and \cref{ShortcutProduct} and  \cref{DrawG,qn,p-centered,non-repetitive,PowerShortcut} imply the following result, which with $g=0$ implies  \cref{kPowerBasic} in the introduction.
%
% \begin{thm}
% \label{PowerGenus}
% For integers $g\geq 0$ and $k,\Delta\geq 1$, let $\ell:= \max\{2g,3\} (2k^4+6k^2) \Delta^{k}$ and $t:= \binom{k+3}{3}-1$. Then for every graph $G$ of Euler genus $g$ and maximum degree $\Delta$,
% \begin{compactitem}
% \item $G^k$ is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$
% \item $G^k$ is $(g, 2k(k+1)\Delta^{k} )$-planar,
% \item $G^k$ has queue-number $\qn(G^k) \leq 3 \ell \cdot 2^t - \ceil{\tfrac{3}{2}\ell}$.
% \item $G^k$ has  $p$-centered chromatic number $\chi_p(G^k) \leq \ell (p+1)\,  \binom{p+t}{t}$,
% \item $G^k$ has  non-repetitive chromatic number $ \pi(G^k) \leq \ell \, 4^{t+1}$.
% \end{compactitem}
% \end{thm}
%
% % \note{DW}{\cref{PowerGenus} has been updated}
%
% This result is the first constant upper bound on the queue-number of bounded powers of graphs with bounded degree and bounded Euler genus.  For every graph $G$, since $G^k$ has maximum degree at most $\Delta^k$, a result of \citet{DJKW16} implies that $\pi(G^k) \leq (1+o(1))\Delta^{2k}$. \cref{PowerGenus} improves upon this bound when $k,g\ll\Delta$.  Similarly, a result of \citet{DFMS21} implies that $\chi_p(G^k)\le 1024p\Delta^{2k}$ and \cref{PowerGenus} improves upon this bound when $p,k,g\ll\Delta$.
%
% \cref{qn,p-centered,non-repetitive,MinorFreeDegree,PowerProduct} imply the following analogous result for powers of graphs in any minor-closed class with bounded maximum degree.
%
% \begin{thm}
% \label{PowerMinor}
% For every graph $X$ there exists an integer $c$ such that for all integers $k,\Delta\geq 1$, if $t:= 2k\Delta^{k}(k^3+3k)\binom{k+c\Delta}{c\Delta}-1$ and $G$ is an $X$-minor-free graph with maximum degree $\Delta$, then:
% \begin{compactitem}
% 	\item $G^k$ is contained in $H\boxtimes P$ for some graph $H$ with treewidth $t$ and for some path $P$.
% 	\item $G^k$ has queue-number at most $3\cdot 2^t-2$,
% 	\item $G^k$ has $p$-centered chromatic number at most $(p+1)\binom{p+t}{t}$.
% \end{compactitem}
% \end{thm}

\subsection{$k$-Nearest-Neighbour Graphs}
\label{last_example}

\cref{nearest-neighbour} and \cref{kPlanarProduct} immediately imply

\begin{thm}
  Every $k$-nearest neighbour graph is a subgraph of $H\boxtimes P\boxtimes K_{\ell}$ for some graph $H$ of treewidth $O(k^6)$ and some $\ell\in O(k^4)$.
\end{thm}

\section{Framed Graphs, $1$-Planar Graphs, and Map Graphs}

% Next we consider
%
% %%%%%%%%%%%%%%%%%%%
% \subsection{Framed Graphs}

% \note{PM}{I don't like this structure anymore.  The proof of \cref{kPlanarProduct} ($k$-planar) refers to the innards of the proof of \cref{ShortcutPartition} ($(k,d)$-shortcuts).  The proofs should be close to each other.
%
% Let's consider something like this:
% \begin{compactenum}
%   \item Introduction
%   \item Shortcut systems and $k$-planar graphs
%   \item All graphs obtained from shortcut systems or $k$-planar graphs
%   \item $d$-framed graphs, $1$-planar graphs, and map graphs
%   \item All applications (colouring, labelling, etc).
% \end{compactenum}
% % (Possibly with the order of 2 and 3 reversed.)
% }

@@

@@

%
% \note{PM}{Use restateable here...}
%
% \begin{thm}
% \label{1-planar}
% Every 1-planar graph is contained in $H\boxtimes P\boxtimes K_7$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
% \end{thm}

@@

\cref{1_planar_product} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma. Similar results appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

@@

\subsection{Universal Graphs}

@@

\section{Orphaned Theorems}

@@

\section{Orphaned Section}

@@

\section{Orphaned Subsection}
